All State Purchase Competition
====

13/4/2014
---
Here we initially analyse the data to begin to understand it. 

The data was first eyeballed in csv format to understand its layout. The problem boils down to predict what options the customers eventually purchases based on their shopping history in order to speed up the sales process.

i.e. 
Predict what options a person settles for and for how much.

The training set contains complete history and the final purchased option
The test sets contains a truncated history. You are not told that the person purchases, i.e. the training set could include people that do not purchase but are just browsing?


Load the data
---

```{r Load the Data}
##Use the data table package, 
##The structure is much quicker than dataframes
##We can also perform grouped calcs very concisely
library(data.table)
data <-  fread("./data/train.csv")
no.observations <- nrow(data)
````

````{r Get the structure of the data}
##Structure of the data
##Number of rows
nrow(data)
##Number of columns
ncol(data)
##Other Summaries
summary(data)
str(data)
````

Lets start with some analysis to understand the data.
This will include checks for missing values or impossible values in each column

customer_ID - A unique identifier for the customer 
---
````{r Customer ID}
setnames(data, "customer_ID","customer")
no.customers <- length(unique(data$customer))
##Number of customers included in training set
no.customers
````
shopping_pt - Unique identifier for the shopping point of a given customer
---
````{r Shopping Point}
setnames(data, "shopping_pt","shopping.pt")
setkey(data, "customer", "shopping.pt")
````
````{r Shopping Point Graphs, fig.width=7, fig.height=6}
hist(data$shopping.pt, main="Distribution of Shopping Points")
summary(data$shopping.pt)
````
The shopping points go from 1 to 6, with mean 4.22, ie. it takes on average 4.22 shopping points to get a hit. Its worth noting some of these shopping points appear to be duplicates.

record_type - 0=shopping point, 1=purchase point
---
````{r Record Type,results='hide'}
data[,record.type:=record_type]
##Create a descriptive factor
set(data,j="record.type",value=factor(data[["record.type"]], levels=c(0,1),labels=c("Quote", "Purchase")))
````

quick check does each customer buy something?
````{r Check Record for each Customer?}
##Use record_type for pivot
sum(data$record_type) == no.customers
````


day - Day of the week (0-6, 0=Monday)
---
````{r Day of Week, results='hide'}
data[,wday:=day]
##Create a descriptive factor
set(data,j="wday",value=factor(data[["wday"]], levels=c(0:6),labels=c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))
````

````{r Number of Shopping Points per day}
##Number of Shopping Points per day
summary(data[,"wday",with=F])
````

````{r Information about Customers}
##Total Number of Customers
no.customers
##Completed purchases Split by day
##Use record_type for pivot
data[,sum(record_type),by=wday]
##Transactions by day
library(plyr)
ddply(data, .(record_type, wday), summarise, no.transactions=length(record_type))
````

time - Time of day (HH:MM)
---
````{r Time of Day, results='hide'}
##Extract the hour and minute of each transaction
data[,hour:=as.numeric(substr(time,1,2))]
data[,min:=as.numeric(substr(time,4,5))]
````
````{r Summary of Times of day}
summary(data[,"hour", with=F])
summary(data[,"min", with=F])
````

state - State where shopping point occurred
---
````{r State of Transaction, results='hide'}
data[,location.state:=state]
##quick check does each customer buy something?
##Create a descriptive factor
set(data,j="location.state",value=factor(data[["location.state"]]))
````
location - Location ID where shopping point occurred
---
````{r Location of Transaction,results='hide'}
data[,location.id:=location]
##quick check does each customer buy something?
##Create a descriptive factor
set(data,j="location.id",value=factor(data[["location.id"]]))
````
````{r Structure of ids}
##Is the id unique per state or overall?
summary.location.id.state <- ddply(data, .(location.id, location.state), summarise,count=length(location.id))
summary.location.id <- as.factor(summary.location.id.state$location.id)
##table values > 1 mean that the id is not unique,
##i.e. different purchases from different states can have the same id
max(table(summary.location.id))
min(table(summary.location.id))
````

group_size - How many people will be covered under the policy (1, 2, 3 or 4)
---
````{r Group Size, results='hide'}
data[,group.size:=as.factor(group_size)]
````
homeowner - Whether the customer owns a home or not (0=no, 1=yes)
---
````{r Homeowner, results='hide'}
data[,homeowner.ind:=factor(homeowner, levels=c(0,1),labels=c("No","Yes"))]
````

car_age - Age of the customer’s car
---
````{r Car Age, results='hide'}
data[,car.age:=as.factor(car_age)]
````
car_value - How valuable was the customer’s car when new
---
````{r Car Value, results='hide'}
data[,car.value:=as.factor(car_value)]
````
risk_factor - An ordinal assessment of how risky the customer is (1, 2, 3, 4)
---
````{r Risk Factor, results='hide'}
data[,risk.factor:=   factor(risk_factor, ordered=T)]
````
age_oldest - Age of the oldest person in customer's group
---
````{r Oldest Age, results='hide'}
data[,age.oldest:=as.factor(age_oldest)]
````
````{r Summary of oldest Age}
##Summary of Oldest Age
summary(data$age.oldest)
hist(data$age.oldest)
````
age_youngest - Age of the youngest person in customer’s group
---
````{r Youngest Age, results='hide'}
data[,age.youngest:=as.factor(age_youngest)]
````
````{r Summary of Youngest Age}
##Summary of Youngest Age
summary(data$age.youngest)
hist(data$youngest.age)
````

married_couple - Does the customer group contain a married couple (0=no, 1=yes)
---
````{r Married Couple, results='hide'}
data[,married.couple:=factor(married_couple, levels=c(0,1), labels=c("Yes","No"))]
````
C_previous - What the customer formerly had or currently has for product option C (0=nothing, 1, 2, 3,4)
---
````{r Previous Product, results='hide'}
data[,c.previous:=C_previous]
````
duration_previous - how long (in years) the customer was covered by their previous issuer
---
````{r Previous Duration, results='hide'}
data[,duration.previous:=as.numeric(duration_previous)]
````
A,B,C,D,E,F,G - the coverage options
---
*A - 0,1,2
*B - 0,1
*C - 1,2,3,4
*D - 1,2,3
*E - 0,1
*F - 0,1,2,3
*G - 1,2,3,4
Formatting the options should ensure no incorrect values and will ensure models fitted use
an appropiate design matrix
````{r Options,results='hide'}
data[,option.a:=factor(A, ordered=T, levels=c(0:2))]
data[,option.b:=factor(B, ordered=T, levels=c(0:1))]
data[,option.c:=factor(C, ordered=T, levels=c(1:4))]
data[,option.d:=factor(D, ordered=T, levels=c(1:3))]
data[,option.e:=factor(E, ordered=T, levels=c(0:1))]
data[,option.f:=factor(F, ordered=T, levels=c(0:3))]
data[,option.g:=factor(G, ordered=T, levels=c(1:4))]
````
cost - cost of the quoted coverage options
---
````{r Cost, fig.width=7, fig.height=6}
summary(data[,"cost", with=F])
hist(data$cost)
plot(density((data$cost)))
````
Layout of final dataset
---
Here we rearrange cols to be in a useful layout and drop unused or raw data cols.
````{r Key Fields to Keep}
key.fields <- c(
                                        # Unique Customer Identifier
  "customer"
  , "location.state"
  , "location.id"
                                        # Unique Transaction
  ,"shopping.pt"
  , "record.type"
                                        # Transaction Time
  , "wday",   "hour", "min"
                                        # Customer Factors/Features
  , "cost", "group.size", "homeowner.ind", "car.age", "car.value", "age.oldest", "age.youngest", "married.couple", "duration.previous", "c.previous", "risk.factor"
                                        # Option values for each Quote/Purchase
  , "option.a", "option.b", "option.c", "option.d", "option.e", "option.f", "option.g"
  )

data.cleaned  <- data[,key.fields, with=F]
setkey(data.cleaned, "customer", "location.state", "location.id", "shopping.pt")
````

Some Further Observations
---
Display all the data
---
````{r Further Observations}
head(data.cleaned, 9)
````
Why does this customer have multiple shopping points at the same time with the same factors?
---
These could be just duplicate records?
We should probably remove these for analysis then recalculate the shopping point numbers, or better to rely on durations of how long a customer spends looking at products

Look at variation in cost per customer
````{r Cost Per Customer, results='hide'}
data.cleaned[,cost.mean:=mean(cost), by=customer]
data.cleaned[,cost.var:=cost-cost.mean]
````

````{r Cost Per Customer (Graph), fig.width=7, fig.height=6}
library(ggplot2)
ggplot(data.cleaned, aes(x=cost.var, y=record.type))+geom_point()+geom_jitter(aes(y=record.type))
````

Look at duration from first to last sale point
Check if some transactions happen on different days
````{r Sales Durations, results='hide'}
data.cleaned[,max(as.numeric(wday))-min(as.numeric(wday)), by=customer]
data.cleaned[J(10000014)]
````
Add an increment column for each day
````{r Delta Days, results='hide'}
day.diff <- c(0,diff(as.numeric(data.cleaned$wday)))
day.diff <- ifelse(day.diff<0,day.diff+7, day.diff)

day.inc <- sapply(2:nrow(data.cleaned), function(x) ifelse(data.cleaned$record.type[x-1]=="Purchase",0,day.diff[x]))

data.cleaned$day.inc <- c(0,day.inc)
````
````{r}
ggplot(data.cleaned, aes(x=day.inc,fill=wday))+geom_histogram()
````
Conclusion: customers who look for products on Sundays buy them that day   

Output Cleaned data
---
write cleaned data to csv for comparrison in spreadsheet form with original
````{r results='hide'}
write.csv(data.cleaned, file="./data/cleandata.csv")
````
